---
title: "MATH 216 Homework 1"
author: "Katherine Hobbs"
output:
  html_document:
    toc: true
    toc_float: true
    collapsed: false
    smooth_scroll: false
---

```{r, echo=FALSE, message=FALSE}
library(ggplot2)
library(dplyr)
# For read_csv() command, which is an improved version of base R's read.csv()
library(readr) 
library(knitr)
library(lubridate)

# Load data sets. Note this assumes this .Rmd files is in the same directory as
# the .csv files.
flights <- read_csv("data/flights.csv") %>% 
  mutate(date=as.Date(date))
weather <- read_csv("data/weather.csv") %>% 
  mutate(date=as.Date(date))
planes <- read_csv("data/planes.csv")
airports <- read_csv("data/airports.csv")
states <- read_csv("data/states.csv")
```





## Admistrative:

Please indicate

* Who you collaborated with: Amanda Hotvedt, Nina Sonneborn, Kyra Gray, 
* Roughly how much time you spent on this HW so far: ~11 hours
* The URL of the RPubs published URL [here](http://rpubs.com/khobbs/213398).
* What gave you the most trouble:

Mostly debugging issues and practicing using each function correctly and in the right context.

* Any comments you have:


## Question 1:

Plot a "time series" of the proportion of flights that were delayed by > 30 minutes on each day.  i.e.
 
* the x-axis should be some notion of time
* the y-axis should be the proportion.

Using this plot, indicate describe the
[seasonality](https://en.wikipedia.org/wiki/Seasonality) of when delays over 30
minutes tend to occur.

```{r, echo=FALSE, message=FALSE, fig.width=12, fig.height=6}

flights <- flights %>% 
  mutate(over30 = ifelse(dep_delay>30, TRUE, FALSE)) %>% 
  filter(!is.na(over30)) %>% 
  group_by(date) %>%
  mutate(prop_over30 = mean(over30))   
  
ggplot(data=flights, aes(x=date, y=prop_over30)) + geom_line() + geom_smooth() + xlab("Month") + ylab("Prop.of Delays Over 30 Min")
  
```

The time series plot depicting the proportion of flights per day leaving Houston that are delayed by over 30 minutes indicates that these types of delays occur most frequently in the winter and spring months, specifically January through June. The proportion of delays over 30 minutes dips for the months of August through November. 

It is important to note that flights with missing delay values were not considered in the analysis of the data set. The eliminated values represented roughly 1.3% of the data set, or 2,905 out of 224,591 entries.

## Question 2:

Some people prefer flying on older planes.  Even though they aren't as nice,
they tend to have more room.  Which airlines should these people favor?

```{r, echo=FALSE, fig.width=12, fig.height=6}

combined <- full_join(planes, flights, by="plane") %>% 
  mutate(age = 2016 - as.numeric(year)) %>% 
  select(age, carrier, year, plane)

combined %>% 
  group_by(carrier) %>% 
  filter(!is.na(age)) %>% 
  summarise(avg_age = mean(age), SD= sd(age), "# of planes"= n()) %>%
  arrange(desc(avg_age)) %>% 
  kable(digits=1)

```
Passangers who enjoy flying on older planes should favor Envoy Air, American Airlines, Delta, and US Airways, as these airlines have the oldest planes, on average. Notably, Envoy Air is a wholly owned subsidiary of American Airlines. 

## Question 3:

* What states did Southwest Airlines' **flight paths** tend to fly to?
* What states did Southwest Airlines' **flights** tend to fly to?

For example, Southwest Airlines Flight 60 to Dallas consists of a single flight
path, but since it flew 299 times in 2013, it would be counted as 299 flights.

```{r, echo=FALSE, message=FALSE, fig.width=12, fig.height=6}

#flight paths
SW_flight_paths <- airports %>% 
  select(iata, state) %>% 
  left_join(flights,., by=c("dest"="iata")) %>% 
  filter(carrier == "WN") %>% 
  group_by(state, flight) %>%
  tally() %>% 
  count(state) %>% 
  arrange(desc(nn))

#flights
SW_flights <- airports %>% 
  select(iata, state) %>% 
  left_join(flights,., by=c("dest"="iata")) %>% 
  filter(carrier == "WN") %>% 
  group_by(state) %>% 
  tally() %>% 
  arrange(desc(n))
  

```
Southwest flight paths from Houston tend to fly to other cities in Texas. After texas, Southwest flights tend to fly to Florida, Louisiana, California, and Oklahoma. 

Similarly, Southwest flights from Houston follow the same pattern, flying most frequently to other cities in Texas and then to Florida, Louisiana, California, and Oklahoma. 


## Question 4:

I want to know proportionately what regions (NE, south, west, midwest) each 
carrier flies to/from Houston in the month of July.  Consider the `month()`
function from the `lubridate` package.

```{r, echo=FALSE, message=FALSE, fig.width=12, fig.height=6}

statesplus <- full_join(states, airports, by = "state") %>% 
  select(statesplus, region, iata)

statesplusflights <- left_join(flights, statesplus, by=c("dest"="iata")) %>% 
  select(date, carrier, dest, region) %>%
  mutate(month=month(date)) %>% 
  filter(month==7) %>% 
  group_by(carrier) %>% 
  count(region) %>% 
  filter(!is.na(region))

ggplot(data=statesplusflights, aes(x=carrier, y=n, fill = region)) +
  geom_bar(stat="identity", position="fill") +
  xlab("Carrier") +
  ylab("Proportion of Flights") +
  ggtitle("Proportion of Flights To and From Houston Arriving in Each Region")
  
  


```
